A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-03 latex/main.tex
Updates from Overleaf

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events, for example,
the deletions inherited by at least 10,000 samples
shown in Table~\ref{tab:major_dels}.
% This just says that the deletions are found in sequences of VOCs.
All these deletions are known to occur in major VOCs
\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,
Tegally2022,Li2023indels},
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of
deletion remapping or sequence misalignment (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).
In the ARG, we found that most of the major deletion events
are associated with the origins of major VOCs
(Table~\ref{tab:major_dels}), and
they have been linked previously with the emergence of these VOCs
\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,Tegally2022}.

Using the ARG, we identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
a 6-base deletion at 21,765,
a 3-base deletion at 21,991, and
a 3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021},
which are also caused by major deletion events in the ARG
(Table~\ref{tab:major_dels}).

Detection of deletions and inference of deletion events
are difficult, especially when the dataset is enormous.
Here we used a naive procedure to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 which commonly occur in BA.1 samples and
the other at 11,288 which commonly occur in BA.2 samples.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
because they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions,
for example, the out-of-frame recurrent deletions
in N, ORF1ab, ORF8, and particularly S1-NTD
(e.g., the 2-base deletion at 21,765)
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

@@

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
686 & 9 & ORF1ab / NSP1 & 6436 \\
21991 & 3 & S / S1-NTD & 2088 \\
21765 & 6 & S / S1-NTD & 482 \\
22029 & 6 & S / S1-NTD & 275 \\
22194 & 3 & S / S1-NTD & 200 \\
6513 & 3 & ORF1ab / NSP3 & 181 \\
21765 & 2 & S / S1-NTD & 140 \\
21995 & 1 & S / S1-NTD & 90 \\
21988 & 3 & S / S1-NTD & 78 \\
21987 & 7 & S / S1-NTD & 78 \\
21993 & 3 & S / S1-NTD & 77 \\
21987 & 9 & S / S1-NTD & 70 \\
21994 & 2 & S / S1-NTD & 65 \\
28370 & 1 & N & 61 \\
11292 & 3 & ORF1ab / NSP6 & 48 \\
21990 & 1 & S / S1-NTD & 46 \\
21987 & 4 & S / S1-NTD & 46 \\
21990 & 6 & S / S1-NTD & 45 \\
686 & 7 & ORF1ab / NSP1 & 43 \\
28252 & 2 & ORF8 & 41 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
To reconstruct
a more comprehensive mutational landscape of SARS-CoV-2,
we mapped deletions onto the ARG using post-hoc parsimony.
We identified major deletion events in the ARG,
which are inherited by at least 10,000 samples
(Table~\ref{tab:major_dels}).
% This just says that the deletions are found in sequences of VOCs.
These deletions are known to occur in major VOCs
\cite{RambautAlpha2020,
Stern2021,
RambautOmicron2021,
Tegally2022,
Li2023indels},
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of deletion remapping (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit of Spike (S1-NTD),
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).
In the ARG, we found that most of these major deletion events
are associated with the origins of major VOCs
(Table~\ref{tab:major_dels}), and
they have been linked previously to the emergence of these VOCs
\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,Tegally2022}.

The ARG enales large-scale detection of recurrent deletions.
Using the ARG, we identified highly recurrent deletions,
for example, the top five deletions
in the number of occurrences in the ARG
(Table~\ref{tab:rec_dels}).
These include known recurrent deletions, for example,
a 6-base deletion at 21,765 \cite{Meng2021} and
a 3-base deletion at 21,991 \cite{McCarthy2021} in S1-NTD,
which are also caused by major deletion events in the ARG
(Table~\ref{tab:major_dels}).

Detection of deletions and inference of deletion events
remain challenging, especially when the dataset is enormous.
Here we took a naive approach to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 and the other at 11,288,
which are commonly found in BA.1 and BA.2 samples, respectively.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
as they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions, for example,
the out-of-frame 4-base recurrent deletion at 21,767 in S1-NTD
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

@@

\begin{table}
\centering
\caption{
Recurrent deletions.
These deletions occur at least 5 times in the ARG and
are found in at least 1,000 samples.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Occurrences & Samples\\
\midrule
686 & 9 & ORF1ab / NSP1 & 6436 & 43354\\
21991 & 3 & S / S1-NTD & 2088 & 309629\\
21765 & 6 & S / S1-NTD & 482 & 820194\\
22029 & 6 & S / S1-NTD & 275 & 1107941\\
22194 & 3 & S / S1-NTD & 200 & 310349\\
6513 & 3 & ORF1ab / NSP3 & 181 & 341380\\
21987 & 9 & S / S1-NTD & 70 & 338633\\
28248 & 6 & ORF8 & 28 & 1113392\\
28271 & 1 & non-coding & 24 & 1331409\\
11288 & 9 & ORF1ab / NSP6 & 19 & 843291\\
11283 & 9 & ORF1ab / NSP6 & 13 & 339810\\
21767 & 4 & S / S1-NTD & 10 & 2144\\
28362 & 9 & N & 7 & 850483\\
21633 & 9 & S / S1-NTD & 7 & 531973\\
\bottomrule
\end{tabular}
\end{table}
